Automated assessment of residual plots with computer vision models

### 2024-02-12 paper/paper.tex
update paper

\caption{\label{tab:performance-sub}The training and test performance of the $32 \times 32$ model presented with different model violations.  The best metrics are colored in red.}
\centering
\resizebox{\linewidth}{!}{
\begin{tabular}[t]{lr>{}l>{}l>{}l>{}l}
\toprule
Violations & \#samples & RMSE & $R^2$ & MAE & Huber loss\\
\midrule
\addlinespace[0.3em]
\multicolumn{6}{l}{\textbf{Training set}}\\
\hspace{1em}no violations & 1546 & \textcolor{black}{1.149} & \textcolor{black}{} & \textcolor{black}{1.080} & \textcolor{black}{0.592}\\
\hspace{1em}non-linearity & 22184 & \textcolor{black}{0.625} & \textcolor{black}{0.851} & \textcolor{black}{0.474} & \textcolor{black}{0.177}\\
\hspace{1em}heteroskedasticity & 10826 & \textcolor{black}{0.528} & \textcolor{black}{0.942} & \textcolor{black}{0.407} & \textcolor{black}{0.133}\\
\hspace{1em}non-linearity + heteroskedasticity & 10221 & \textcolor{black}{0.593} & \textcolor{black}{0.920} & \textcolor{black}{0.439} & \textcolor{black}{0.158}\\
\hspace{1em}non-normality & 10797 & \textcolor{red}{0.279} & \textcolor{red}{0.955} & \textcolor{red}{0.177} & \textcolor{red}{0.038}\\
\hspace{1em}non-linearity + non-normality & 8835 & \textcolor{black}{0.472} & \textcolor{black}{0.923} & \textcolor{black}{0.274} & \textcolor{black}{0.090}\\
\hspace{1em}heteroskedasticity + non-normality & 7870 & \textcolor{black}{0.354} & \textcolor{black}{0.905} & \textcolor{black}{0.211} & \textcolor{black}{0.052}\\
\hspace{1em}non-linearity + heteroskedasticity + non-normality & 7721 & \textcolor{black}{0.432} & \textcolor{black}{0.874} & \textcolor{black}{0.261} & \textcolor{black}{0.078}\\
\addlinespace[0.3em]
\multicolumn{6}{l}{\textbf{Test set}}\\
\hspace{1em}no violations & 155 & \textcolor{black}{1.267} & \textcolor{black}{} & \textcolor{black}{1.174} & \textcolor{black}{0.685}\\
\hspace{1em}non-linearity & 2218 & \textcolor{black}{0.787} & \textcolor{black}{0.771} & \textcolor{black}{0.578} & \textcolor{black}{0.257}\\
\hspace{1em}heteroskedasticity & 1067 & \textcolor{black}{0.602} & \textcolor{black}{0.922} & \textcolor{black}{0.461} & \textcolor{black}{0.170}\\
\hspace{1em}non-linearity + heteroskedasticity & 985 & \textcolor{black}{0.751} & \textcolor{black}{0.868} & \textcolor{black}{0.541} & \textcolor{black}{0.236}\\
\hspace{1em}non-normality & 1111 & \textcolor{red}{0.320} & \textcolor{red}{0.942} & \textcolor{red}{0.194} & \textcolor{red}{0.049}\\
\hspace{1em}non-linearity + non-normality & 928 & \textcolor{black}{0.600} & \textcolor{black}{0.879} & \textcolor{black}{0.335} & \textcolor{black}{0.133}\\
\hspace{1em}heteroskedasticity + non-normality & 819 & \textcolor{black}{0.489} & \textcolor{black}{0.832} & \textcolor{black}{0.267} & \textcolor{black}{0.093}\\
\hspace{1em}non-linearity + heteroskedasticity + non-normality & 717 & \textcolor{black}{0.620} & \textcolor{black}{0.762} & \textcolor{black}{0.339} & \textcolor{black}{0.140}\\
\bottomrule
\end{tabular}}
\end{table}

@@

In \citet{li2023plot}, the residual plots are simulated from a data
generating process which is a special case of Equation
\ref{eq:data-sim}. The main characteristic is the model violations are
introduced separately, meaning non-linearity and heteroskedasticity will
not co-exist in one lineup but assigned uniformly to all lineups.
Additionally, non-normality and multiple predictors are not considered
in their experimental design.

@@

For each lineup used in \citet{li2023plot}, there are one actual
residual plot and \(19\) null plots. While the distance \(D\) for the
actual residual plot depends on the underlying data generating process,
the distance \(D\) for the null plots are zeros. We have used our
optimized computer vision model to predict \(\hat{D}\) for both the
actual residual plots and the null plots.

The performance metrics of \(\hat{D}\) for actual residual plots are
provided in Table \ref{tab:experiment-performance}. It can be observed
that all the performance metrics are slightly worse than the metrics
evaluated on the test data. Still, the mean absolute error remains at a
low level, and the linearly correlation between the prediction and the
true value is still very high. Lineups with non-linearity issues are
more difficult to be predicted than those with heteroskedasticity
issues.

For null plots, a histogram of the prediction is provided in Figure
\ref{fig:hist-null-human}. In most cases, \(\hat{D}\) will be predicted
at around \(1\). And as \(\hat{D}\) increases, the number of predictions
decreases. There are some outliers with \(\hat{D} > 5\).

@@

\begin{figure}

@@

\begin{figure}

{\centering \includegraphics[width=1\linewidth]{paper_files/figure-latex/unnamed-chunk-4-1}

@@

\caption{A weighted detection rate vs $\delta$-differnence plot.}\label{fig:unnamed-chunk-4}
\end{figure}

\begin{figure}

{\centering \includegraphics[width=1\linewidth]{paper_files/figure-latex/unnamed-chunk-5-1}

@@

\caption{A weighted detection rate vs $\gamma$-number of extreme nulls plot.}\label{fig:unnamed-chunk-5}
\end{figure}

\begin{figure}

{\centering \includegraphics[width=1\linewidth]{paper_files/figure-latex/unnamed-chunk-6-1}

@@

\caption{A weighted detection rate vs $\delta$-differnence plot for lineups evaluated by 11 people.}\label{fig:unnamed-chunk-6}
\end{figure}

\begin{figure}

{\centering \includegraphics[width=1\linewidth]{paper_files/figure-latex/unnamed-chunk-7-1}

\caption{A weighted detection rate vs $\gamma$-number of extreme nulls plot for lineups evaluated by 11 people.}\label{fig:unnamed-chunk-7}
\end{figure}

\hypertarget{when-the-model-works}{%
\subsection{When the model works}\label{when-the-model-works}}

\begin{itemize}
\tightlist
\item
  simple examples (non-linearity, heteroskedasticity, \ldots)
\item
  datasaurus
\end{itemize}

\hypertarget{when-the-model-does-not-work}{%
\subsection{When the model does not
work}\label{when-the-model-does-not-work}}

\begin{itemize}
\tightlist
\item
  human detect but model does not
\item
  cartoon residuals?
\end{itemize}

\hypertarget{workflow-how-one-use-this-model-small-showcase}{%
\subsection{Workflow: how one use this model? (small
showcase)}\label{workflow-how-one-use-this-model-small-showcase}}

\hypertarget{dicussion}{%
\section{Dicussion}\label{dicussion}}

There are other kinds of residual departures like autocorrelation that
are not considered in this study. The primary goal of this study is to
establish a new way of evaluating residual plot and conducting visual
test with computer vision models. Building computer vision models for
other model violations and other types of diagnostic plots could be
future directions of this field.

>>>>

\caption{\label{tab:performance-sub}The training and test performance of the $32 \times 32$ model presented with different model violations.  The best metrics are colored in red.}
\centering
\resizebox{\linewidth}{!}{
\begin{tabular}[t]{ll>{}l>{}l>{}l>{}l}
\toprule
Violations & \#samples & RMSE & $R^2$ & MAE & Huber loss\\
\midrule
\addlinespace[0.3em]
\multicolumn{6}{l}{\textbf{Training set}}\\
\hspace{1em}no violations & 1546 & \textcolor{black}{1.149} & \textcolor{black}{} & \textcolor{black}{1.080} & \textcolor{black}{0.592}\\
\hspace{1em}non-linearity & 22184 & \textcolor{black}{0.625} & \textcolor{black}{0.851} & \textcolor{black}{0.474} & \textcolor{black}{0.177}\\
\hspace{1em}heteroskedasticity & 10826 & \textcolor{black}{0.528} & \textcolor{black}{0.942} & \textcolor{black}{0.407} & \textcolor{black}{0.133}\\
\hspace{1em}non-linearity + heteroskedasticity & 10221 & \textcolor{black}{0.593} & \textcolor{black}{0.920} & \textcolor{black}{0.439} & \textcolor{black}{0.158}\\
\hspace{1em}non-normality & 10797 & \textcolor{red}{0.279} & \textcolor{red}{0.955} & \textcolor{red}{0.177} & \textcolor{red}{0.038}\\
\hspace{1em}non-linearity + non-normality & 8835 & \textcolor{black}{0.472} & \textcolor{black}{0.923} & \textcolor{black}{0.274} & \textcolor{black}{0.090}\\
\hspace{1em}heteroskedasticity + non-normality & 7870 & \textcolor{black}{0.354} & \textcolor{black}{0.905} & \textcolor{black}{0.211} & \textcolor{black}{0.052}\\
\hspace{1em}non-linearity + heteroskedasticity + non-normality & 7721 & \textcolor{black}{0.432} & \textcolor{black}{0.874} & \textcolor{black}{0.261} & \textcolor{black}{0.078}\\
\addlinespace[0.3em]
\multicolumn{6}{l}{\textbf{Test set}}\\
\hspace{1em}no violations & 155 & \textcolor{black}{1.267} & \textcolor{black}{} & \textcolor{black}{1.174} & \textcolor{black}{0.685}\\
\hspace{1em}non-linearity & 2218 & \textcolor{black}{0.787} & \textcolor{black}{0.771} & \textcolor{black}{0.578} & \textcolor{black}{0.257}\\
\hspace{1em}heteroskedasticity & 1067 & \textcolor{black}{0.602} & \textcolor{black}{0.922} & \textcolor{black}{0.461} & \textcolor{black}{0.170}\\
\hspace{1em}non-linearity + heteroskedasticity & 985 & \textcolor{black}{0.751} & \textcolor{black}{0.868} & \textcolor{black}{0.541} & \textcolor{black}{0.236}\\
\hspace{1em}non-normality & 1111 & \textcolor{red}{0.320} & \textcolor{red}{0.942} & \textcolor{red}{0.194} & \textcolor{red}{0.049}\\
\hspace{1em}non-linearity + non-normality & 928 & \textcolor{black}{0.600} & \textcolor{black}{0.879} & \textcolor{black}{0.335} & \textcolor{black}{0.133}\\
\hspace{1em}heteroskedasticity + non-normality & 819 & \textcolor{black}{0.489} & \textcolor{black}{0.832} & \textcolor{black}{0.267} & \textcolor{black}{0.093}\\
\hspace{1em}non-linearity + heteroskedasticity + non-normality & 717 & \textcolor{black}{0.620} & \textcolor{black}{0.762} & \textcolor{black}{0.339} & \textcolor{black}{0.140}\\
\bottomrule
\end{tabular}}
\end{table}

@@

In \citet{li2023plot}, the residual plots are simulated from a data
generating process which is a special case of Equation
\ref{eq:data-sim}. The main characteristic is the model violations are
introduced separately, meaning non-linearity and heteroskedasticity will
not co-exist in one lineup but assigned uniformly to all lineups.
Additionally, non-normality and multiple predictors are not considered
in the experimental design.

@@

@@

For each lineup used in \citet{li2023plot}, there is one actual residual
plot and 19 null plots. While the distance \(D\) for the actual residual
plot depends on the underlying data generating process, the distance
\(D\) for the null plots is zero. We have used our optimized computer
vision model to predict \(\hat{D}\) for both the actual residual plots
and the null plots.

The performance metrics of \(\hat{D}\) for actual residual plots are
outlined in Table \ref{tab:experiment-performance}. It's notable that
all performance metrics are slightly worse than those evaluated on the
test data. Nevertheless, the mean absolute error remains at a low level,
and the linear correlation between the prediction and the true value
remains very high. Lineups with non-linearity issues are more
challenging to predict than those with heteroskedasticity issues.

For null plots, a histogram of the predictions is provided in Figure
\ref{fig:hist-null-human}. In most cases, \(\hat{D}\) is predicted to be
around \(1\). As \(\hat{D}\) increases, the number of predictions
decreases. The 90\% and 95\% sample quantiles are approximately at
\(2\), and the 99\% sample quantile is around \(2.5\). These sample
quantiles can serve as global critical values for determining whether a
residual plot needs to be rejected. We have also computed the 95\%
sample quantiles by assessing 200 null plots for each fitted model used
in the experiment. The difference between decisions based on these two
types of critical values is minor, with decisions aligning 93.8\% of the
time. Therefore, we will be using the 95\% sample quantiles obtained
from each fitted model in the following analysis.

Figure \ref{fig:human-mosaic} and Table \ref{tab:human-table} provide
insights into the agreement between decisions made by the computer
vision model and visual tests conducted by humans. It can be observed
that visual tests are less sensitive than the computer vision model, as
they reject almost only when the computer vision model does. The
agreement rates between visual tests and the computer vision model are
67.96\% and 66.84\% for residual plots containing heteroskedasticity and
non-linearity patterns, respectively.

When plotting the decision against the distance, as depicted in Figure
\ref{fig:power}, we observe that for the computer vision model, when the
distance \(D > 3\), almost all residual plots will be rejected. However,
for visual tests conducted by humans, the threshold is higher at
\(D > 5\). Visual tests also demonstrate low sensitivity to residual
plots with small distances compared to the computer vision model.

Figure \ref{fig:delta} displays the scatter plot of the weighted
detection rate vs the \(\delta\)-difference. In the experiment conducted
in \citet{li2023plot}, participants were allowed to make multiple
selections for a lineup. The weighted detection rate is computed by
assigning weights to each detection. If the participant selects zero
plots, the corresponding weight is 0.05; otherwise, the weight is 1
divided by the number of selections. The \(\delta\)-difference is
defined by \citet{chowdhury2018measuring} as

\begin{equation}
\delta = \hat{D} - \underset{j}{max}\left(\hat{D}_{null}^{(j)}\right) \quad \text{for}~j = 1,...,m-1,
\end{equation}

\noindent where \(\hat{D}_{null}^{(j)}\) is the approximated distance
for the \(j\)-th null plot, and \(m\) is the number of plots in a
lineup.

Figure \ref{fig:delta} indicates that the weighted detection rate
increases as the \(\delta\)-difference increases, particularly when the
\(\delta\)-difference is greater than zero. A negative
\(\delta\)-difference suggests that there is at least one null plot in
the lineup with a stronger visual signal than the actual residual plot.
In some instances, the weighted detection rate is close to one, yet the
\(\delta\)-difference is negative. This discrepancy implies that the
distance measure, or the approximated distance, may not perfectly
reflect actual human behavior.

@@

\caption{\label{tab:human-table}Summary of decisions made by the visual test conducted by human and decisions made by the computer vision model.}
\centering
\begin{tabular}[t]{llll}
\toprule
Violations & \#Samples & \#Agreements & Agreement rate\\
\midrule
heteroskedasticity & 540 & 367 & 0.6796\\
non-linearity & 576 & 385 & 0.6684\\
\bottomrule
\end{tabular}
\end{table}

{\centering \includegraphics[width=1\linewidth]{paper_files/figure-latex/human-mosaic-1}

@@

\caption{Rejection rate ($p$-value $\leq0.05$) of visual test conditional on the computer vision model decision on non-linearity (left) and heteroskedasticity (right) lineups displayed using a mosaic plot. The visual test rejects less frequently than the computer vision model, and (almost) only rejects when the computer vision model does.}\label{fig:human-mosaic}
\end{figure}

{\centering \includegraphics[width=1\linewidth]{paper_files/figure-latex/power-1}

@@

\caption{Comparison of decisions made by visual tests and the computer vision model. The data points are fitted with logistic regression models with no intercept but an offset equals to $log(0.05/0.95)$.}\label{fig:power}
\end{figure}

{\centering \includegraphics[width=1\linewidth]{paper_files/figure-latex/delta-1}

@@

\caption{A weighted detection rate vs $\delta$-differnence plot.}\label{fig:delta}
\end{figure}

\hypertarget{data-examples}{%
\subsection{Data examples}\label{data-examples}}

\hypertarget{boston-housing}{%
\subsubsection{Boston Housing}\label{boston-housing}}

\begin{center}\includegraphics[width=1\linewidth]{paper_files/figure-latex/unnamed-chunk-8-1} \end{center}

\hypertarget{advertising}{%
\subsubsection{Advertising}\label{advertising}}

\begin{center}\includegraphics[width=1\linewidth]{paper_files/figure-latex/unnamed-chunk-12-1} \end{center}

\hypertarget{wine-quality}{%
\subsubsection{Wine Quality}\label{wine-quality}}

\begin{center}\includegraphics[width=1\linewidth]{paper_files/figure-latex/unnamed-chunk-16-1} \end{center}

\hypertarget{datasaurus}{%
\subsubsection{Datasaurus}\label{datasaurus}}

\begin{center}\includegraphics[width=1\linewidth]{paper_files/figure-latex/unnamed-chunk-20-1} \end{center}
